Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-16 cordillera-cycle.tex
Typeset: no-break space after 'A' and 'a'.

      During the last glacial cycle, glaciers and ice caps of the North
      American Cordillera have been more extensive than today. At the Last
      Glacial Maximum (LGM), a continuous blanket of ice, the Cordilleran
      ice sheet \citep{Dawson.1888}, stretched from the Alaska Range in the
      north to the North Cascades in the south (Fig.~\ref{fig:locmap}). In
      addition, it extended offshore, where it calved into the Pacific
      Ocean, and merged with the western margin of its much larger
      neighbour, the Laurentide ice sheet, east of the Rocky Mountains.

      More than a century of exploration and geological investigation of the
      Cordilleran mountains have led to many observations in support of the
      former ice sheet \citep{Jackson.Clague.1991}. Despite the lack of
      documented end moraines offshore, in the zone of confluence with the
      Laurentide ice sheet, and in areas swept by the Missoula floods
      \citep{Carrara.etal.1996}, moraines that demarcate the northern and
      south-western margins provide key constraints that allow reasonable
      reconstructions of maximum ice sheet extents
      (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
      \citealp{Duk-Rodkin.1999}; \citealp{Booth.etal.2003};
      \citealp{Dyke.2004}). As indicated by field evidence from radiocarbon
      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
      \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
      bedrock deformation in response to former ice loads
      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
      records \citep{Cosma.etal.2008, Davies.etal.2011}, the LGM Cordilleran
      ice sheet extent was short-lived. However, former ice thicknesses and,
      therefore, the ice sheet's contribution to the LGM sea level lowstand
      \citep{Carlson.Clark.2012, Clark.Mix.2002} remain uncertain.

@@

      In contrast, evidence for the deglaciation history of the Cordilleran
      ice sheet since the LGM is considerable, albeit mostly at a regional
      scale. Geomorphological evidence from south-central British Columbia
      indicates a rapid deglaciation, including an early emergence of
      elevated areas while thin, stagnant ice still covered the surrounding
      lowlands \citep{Fulton.1967, Fulton.1991, Margold.etal.2011,
      Margold.etal.2013a}. This model, although credible, may not apply in
      all areas of the Cordilleran ice sheet \citep{Margold.etal.2013}.
      Although solid evidence for late-glacial glacier re-advances have been
      found in the Coast, Columbia and Rocky mountains
      \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
      Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
      Menounos.etal.2008}, it appears to be sparser than for formerly
      glaciated regions surrounding the North Atlantic
      \citep[e.g.,][]{Sissons.1979, Lundqvist.1987, Ivy-Ochs.etal.1999,
      Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements from
      Gulf of Alaska sediments reveal a climatic evolution highly correlated
      to that of Greenland during this period, including a distinct Late
      Glacial cold reversal between 14.1 and 11.7\,ka
      \citep{Praetorius.Mix.2014}.

      In general, the topographic complexity of the North American
      Cordillera and its effect on glacial history have inhibited the
      reconstruction of ice sheet-wide glacial advance and retreat patterns
      such as those available for the Fennoscandian and Laurentide ice
      sheets \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
      Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}. Here, we
      use a numerical ice sheet model \citep{PISM-authors.2015}, calibrated
      against field-based evidence, to perform a quantitative reconstruction
      of the Cordilleran ice sheet evolution through the last glacial cycle,
      and analyse some of the long-standing questions related to its
      evolution:

@@

      Our palaeo-climate forcing therefore includes spatial temperature and
      precipitation grids derived from a present-day atmospheric reanalysis
      \citep{Mesinger.etal.2006} that includes the steep precipitation
      gradients previously identified as necessary to model the LGM extent
      of the Cordilleran ice sheet in agreement with its geological imprint
      \citep{Seguinot.etal.2014}. To mimic climate evolution through the
      last glacial cycle, these grids are simply supplemented by lapse-rate
      corrections and temperature offset time series. The latter are
      obtained by scaling six different palaeo-temperature reconstructions
      from proxy records around the globe, including two oxygen isotope
      records from Greenland ice cores \citep{Dansgaard.etal.1993,
      Andersen.etal.2004}, two oxygen isotope records from Antarctic ice
      cores \citep{Petit.etal.1999,Jouzel.etal.2007}, and two alkenone
      unsaturation index records from Northwest Pacific ocean sediment cores
      \citep{Herbert.etal.2001}. We then proceed to compare the model output
      to geological evidence and discuss the timing and extent of glaciation
      and the patterns of deglaciation, based on which we use the
      applicability of different records to modelling the history of the
      Cordilleran ice sheet.

@@

      Basal topography is derived from the ETOPO1 combined topography and
      bathymetry dataset with a~resolution of 1\,arc-min
      \citep{Amante.Eakins.2009}. Sea level is lowered as a function of time
      based on the Spectral Mapping Project
      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
      is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
      (Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a
      positive degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate
      forcing is provided by a monthly climatology averaged from 1979 to
      2000 from the North American Regional Reanalysis
      \citep[NARR,][]{Mesinger.etal.2006}, perturbated by time-dependent
      offsets and lapse-rate temperature corrections (Sect.~\ref{sec:atm}).

      Each simulation starts from assumed ice-free conditions at 120000
      years ago (120\,ka), and runs to the present. Our modelling domain of
      1500 by 3000\,km encompasses the entire area covered by the
      Cordilleran ice sheet at the LGM (Fig.~\ref{fig:locmap}). The
      simulations were run on two distinct grids, using a lower horizontal
      resolution of 10\,km, and a higher horizontal resolution of 5\,km.
      These computations were performed on 16 to 128 computing cores at the
      Swedish National Supercomputing Centre.

@@

      Ice sheet dynamics are typically modelled using a combination of
      internal deformation and basal sliding. PISM is a~shallow ice sheet
      model, which implies that the balance of stresses is approximated
      based on their predominant components. The Shallow Shelf Approximation
      (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation
      (SIA) by adding velocity solutions of the two approximations
      \citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.

      Ice rheology depends on temperature and water content through an
      enthalpy formulation \citep{Aschwanden.etal.2012}. Surface air
      temperature derived from the climate forcing (Sect.~\ref{sec:atm})
      provides the upper boundary condition to the ice enthalpy model.
      Temperature is computed subglacially to a~depth of 3\,km, where it is
      conditioned by a lower boundary geothermal heat flux of
      70\,\unit{mW\,m^{-2}}. Although this uniform value does not account
      for the high spatial geothermal variability in the region
      \citep{Blackwell.Richards.2004}, it is, on average, representative of
      available heat flow measurements. In the low-resolution simulations,
      the vertical grid consists of 31~temperature layers in the bedrock and
      up to 51~enthalpy layers in the ice sheet, corresponding to a vertical
      resolution of 100\,m. The high-resolution simulations 61~bedrock
      layers and up to 101~ice layers with a vertical resolution of 50\,m.

      A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
      relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the
      sliding velocity, $\vec{v}_b$ (Table~\ref{tab:params}). The yield
      stress, $\tau_c$, is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
      where cohesion, $c_0$, is assumed to be zero. The friction angle,
      $\phi$, varies from 15 to 45\degree as a piecewise-linear function of
      modern bed elevation, with the lowest value occuring below modern sea
      level (0\,m above sea level, m~a.s.l.) and the highest value occuring
      above the generalised elevation of the highest shorelines
      \citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
      a~weakening of till associated with the presence of marine sediments
      (cf. \citealp{Martin.etal.2011};
      \citealp[supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
      the ice overburden stress, $\rho gh$, and the modelled amount of
      subglacial water, using a formula derived from laboratory experiments
      with till extracted from the base of Ice Stream B in West Antarctica
      \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
      where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void
      ratio and $C_c$ is a measured compressibility coefficient
      (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
      from zero to $W_{max}=2$\,m, a threshold above which additional melt
      water is assumed to drain off instantaneously. Finally, the bedrock
      topography responds to ice load following a bedrock deformation model
      that includes local isostasy, elastic lithosphere flexure and viscous
      astenosphere deformation in an infinite half-space
      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A relatively low viscosity
      value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is used for the
      astenosphere (Table~\ref{tab:params}) in accordance with the results
      from regional glacial isostatic adjustment modelling at the northern
      Cascadia subduction zone \citep{James.etal.2009}.

      Ice shelf calving is computed using a double criterion. First, a
      physically-realistic calving flux is computed based on eigenvalues of
      the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
      Levermann.etal.2012}. This allows floating ice to advance in confined
      embayments, but prevents the formation of extensive ice shelves in the
      open ocean. Second, floating ice thinner than 50\,m is systematically
      calved off. A subgrid scheme by \citet{Albrecht.etal.2011} allows for
      a continuous migration of the calving front. This formulation of
      calving has been applied to the Antarctic ice sheet and has shown to
      produce a realistic calving front positon for many of the present-day
      ice shelves \citep{Martin.etal.2011}.

@@

      The PDD computation accounts for stochastic temperature variations by
      assuming a normal temperature distribution of standard deviation
      $\sigma$ aroung the expected value $T_m$. It is expressed by an
      error-function formulation \citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
      which is numerically approximated using week-long sub-intervals. In
      order to account for the effects of spatial and seasonal variations of
      temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
      from NARR daily temperature values from 1979 to 2000
      \citep{Mesinger.etal.2006}, including variability associated with the
      seasonal cycle (Fig.~\ref{fig:atm}). Degree-day factors for snow and
      ice melt are derived from mass-balance measurements on contemporary
      glaciers from the Coast Mountains and Rocky Mountains in British
      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

      The present-day monthly climatology was computed from near-surface air
      temperature and precipitation rate fields from the NARR, averaged from
      1979 to 2000. Modern climate of the North American Cordillera is
      characterised by strong geographic variations in temperature
      seasonality, timing of the maximum annual precipitation, and daily
      temperature variability (Fig.~\ref{fig:atm}). The use of NARR is
      motivated by the need for an accurate, high-resolution precipitation
      forcing, as identified in a previous sensitivity study
      \citep{Seguinot.etal.2014}.

      Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
      palaeo-temperature proxy records from the Greenland Ice Core Project
      \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice Core
      Project \citep[NGRIP,][]{Andersen.etal.2004}, the European Project for
      Ice Coring in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, the
      Vostok ice core \citep{Petit.etal.1999}, and Ocean Drilling Program
      (ODP) sites 1012 and 1020, both located off the coast of California
      \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the GRIP
      and NGRIP records were calculated from oxygen isotope
      (\chem{\delta^{18}O}) measurements using a quadratic equation
      \citep{Johnsen.etal.1995},
\begin{align}
    {\Delta}T_{TS}(t) = &-11.88 [\chem{\delta^{18}O}(t)
                                -\chem{\delta^{18}O}(0)] \nonumber \\
                        &-0.1925[\chem{\delta^{18}O}(t)^2
                                 -\chem{\delta^{18}O}(0)^2] \,,
\end{align}
      while temperature reconstructions from Antarctic and oceanic cores
      were provided as such. All records were scaled linearly
      (Table~\ref{tab:records}) in order to simulate comparable ice extents
      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
      (Fig.~\ref{fig:lr_maps}).

      Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a
      function of ice surface elevation, $s$, using the NARR surface
      geopotential height invariant field as a reference topography,
      $b_{ref}$:
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
\end{align}

@@

      Despite large differences in the input climate forcing
      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
      features that can be observed across the range of forcing data used.
      In all simulations, modelled ice volumes remain relatively low during
      most of the glacial cycle, except during two major glacial events
      which occur between 61.9 and 56.5\,ka during MIS~4, and between 23.2
      and 16.8\,ka during MIS~2 (Fig.~\ref{fig:lr_ts}, lower panel). An ice
      volume minimum is consistently reached between 53.0 and 41.3\,ka
      during MIS~3. However, the magnitude and precise timing of these three
      events depend significantly on the choice of proxy record used to
      derive a time-dependent climate forcing (Table~\ref{tab:extrema}).

      Simulations forced by the Greenland ice core palaeo-temperature
      records (GRIP, NGRIP) produce the highest variability in modelled ice
      volume throughout the last glacial cycle. In contrast, simulations
      driven by oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
      palaeo-temperature records generally result in lower ice volume
      variability throughout the simulation length, resulting in lower
      modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
      The NGRIP climate forcing is the only one that results in a larger ice
      volume during MIS~4 than during MIS~2.

      While simulations driven by the GRIP and the two Antarctic
      palaeo-temperature records attain a last ice volume maximum between
      19.1 and 16.8\,ka, those informed by the NGRIP and the two oceanic
      palaeo-temperature records attain their maximum ice volumes thousands
      of years earlier. Moreover, the ODP~1012 run yields a rapid
      deglaciation of the modelled area prior to 17\,ka. The ODP~1020
      simulation predicts an early maximum in ice volume at 23.2\,ka,
      followed by slower deglaciation than modelled using the other
      palaeo-temperature records. Finally, whereas model runs forced by the
      Antarctic palaeo-temperature records result in a rapid and
      uninterrupted deglaciation after the LGM, the simulation driven by the
      GRIP palaeo-temperature record also results in a rapid deglaciation
      but in three steps, separated by two periods of ice sheet regrowth
      (Fig.~\ref{fig:lr_ts}).

@@

      In the MIS~3 ice volume minimum reconstructions, a central ice cap
      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
      panels). Although this ice cap is present in all simulations, its
      dimensions depend sensitively on the choice of the applied
      palaeo-temperature record. Modelled ice volume minima spread over a
      wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).

      Modelled ice sheet geometries during the LGM (MIS~2;
      Fig.~\ref{fig:lr_maps}, lower panels) invariably include a ca.
      1500\,km-long central divide above 3000\,m~a.s.l. located along the
      spine of the Rocky Mountains. Although the similarity of modelled ice
      extents is a direct result of the choice of scaling factors applied to
      different palaeo-temperature proxy records (Table~\ref{tab:records}),
      it is interesting to note that modelled maximum ice volumes also fall
      within a tight range of 8.40 to 8.91\,m~s.l.e.
      (Table~\ref{tab:extrema}).

>>>>

      During the last glacial cycle, glaciers and ice caps of the North
      American Cordillera have been more extensive than today. At the Last
      Glacial Maximum (LGM), a~continuous blanket of ice, the Cordilleran
      ice sheet \citep{Dawson.1888}, stretched from the Alaska Range in the
      north to the North Cascades in the south (Fig.~\ref{fig:locmap}). In
      addition, it extended offshore, where it calved into the Pacific
      Ocean, and merged with the western margin of its much larger
      neighbour, the Laurentide ice sheet, east of the Rocky Mountains.

      More than a~century of exploration and geological investigation of the
      Cordilleran mountains have led to many observations in support of the
      former ice sheet \citep{Jackson.Clague.1991}. Despite the lack of
      documented end moraines offshore, in the zone of confluence with the
      Laurentide ice sheet, and in areas swept by the Missoula floods
      \citep{Carrara.etal.1996}, moraines that demarcate the northern and
      south-western margins provide key constraints that allow reasonable
      reconstructions of maximum ice sheet extents
      (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
      \citealp{Duk-Rodkin.1999}; \citealp{Booth.etal.2003};
      \citealp{Dyke.2004}). As indicated by field evidence from radiocarbon
      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
      \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
      bedrock deformation in response to former ice loads
      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
      records \citep{Cosma.etal.2008, Davies.etal.2011}, the LGM Cordilleran
      ice sheet extent was short-lived. However, former ice thicknesses and,
      therefore, the ice sheet's contribution to the LGM sea level lowstand
      \citep{Carlson.Clark.2012, Clark.Mix.2002} remain uncertain.

@@

      In contrast, evidence for the deglaciation history of the Cordilleran
      ice sheet since the LGM is considerable, albeit mostly at a~regional
      scale. Geomorphological evidence from south-central British Columbia
      indicates a~rapid deglaciation, including an early emergence of
      elevated areas while thin, stagnant ice still covered the surrounding
      lowlands \citep{Fulton.1967, Fulton.1991, Margold.etal.2011,
      Margold.etal.2013a}. This model, although credible, may not apply in
      all areas of the Cordilleran ice sheet \citep{Margold.etal.2013}.
      Although solid evidence for late-glacial glacier re-advances have been
      found in the Coast, Columbia and Rocky mountains
      \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
      Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
      Menounos.etal.2008}, it appears to be sparser than for formerly
      glaciated regions surrounding the North Atlantic
      \citep[e.g.,][]{Sissons.1979, Lundqvist.1987, Ivy-Ochs.etal.1999,
      Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements from
      Gulf of Alaska sediments reveal a~climatic evolution highly correlated
      to that of Greenland during this period, including a~distinct Late
      Glacial cold reversal between 14.1 and 11.7\,ka
      \citep{Praetorius.Mix.2014}.

      In general, the topographic complexity of the North American
      Cordillera and its effect on glacial history have inhibited the
      reconstruction of ice sheet-wide glacial advance and retreat patterns
      such as those available for the Fennoscandian and Laurentide ice
      sheets \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
      Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}. Here, we
      use a~numerical ice sheet model \citep{PISM-authors.2015}, calibrated
      against field-based evidence, to perform a~quantitative reconstruction
      of the Cordilleran ice sheet evolution through the last glacial cycle,
      and analyse some of the long-standing questions related to its
      evolution:

@@

      Our palaeo-climate forcing therefore includes spatial temperature and
      precipitation grids derived from a~present-day atmospheric reanalysis
      \citep{Mesinger.etal.2006} that includes the steep precipitation
      gradients previously identified as necessary to model the LGM extent
      of the Cordilleran ice sheet in agreement with its geological imprint
      \citep{Seguinot.etal.2014}. To mimic climate evolution through the
      last glacial cycle, these grids are simply supplemented by lapse-rate
      corrections and temperature offset time series. The latter are
      obtained by scaling six different palaeo-temperature reconstructions
      from proxy records around the globe, including two oxygen isotope
      records from Greenland ice cores \citep{Dansgaard.etal.1993,
      Andersen.etal.2004}, two oxygen isotope records from Antarctic ice
      cores \citep{Petit.etal.1999,Jouzel.etal.2007}, and two alkenone
      unsaturation index records from Northwest Pacific ocean sediment cores
      \citep{Herbert.etal.2001}. We then proceed to compare the model output
      to geological evidence and discuss the timing and extent of glaciation
      and the patterns of deglaciation, based on which we use the
      applicability of different records to modelling the history of the
      Cordilleran ice sheet.

@@

      Basal topography is derived from the ETOPO1 combined topography and
      bathymetry dataset with a~resolution of 1\,arc-min
      \citep{Amante.Eakins.2009}. Sea level is lowered as a~function of time
      based on the Spectral Mapping Project
      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
      is applied as a~constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
      (Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a
      positive degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate
      forcing is provided by a~monthly climatology averaged from 1979 to
      2000 from the North American Regional Reanalysis
      \citep[NARR,][]{Mesinger.etal.2006}, perturbated by time-dependent
      offsets and lapse-rate temperature corrections (Sect.~\ref{sec:atm}).

      Each simulation starts from assumed ice-free conditions at 120000
      years ago (120\,ka), and runs to the present. Our modelling domain of
      1500 by 3000\,km encompasses the entire area covered by the
      Cordilleran ice sheet at the LGM (Fig.~\ref{fig:locmap}). The
      simulations were run on two distinct grids, using a~lower horizontal
      resolution of 10\,km, and a~higher horizontal resolution of 5\,km.
      These computations were performed on 16 to 128 computing cores at the
      Swedish National Supercomputing Centre.

@@

      Ice sheet dynamics are typically modelled using a~combination of
      internal deformation and basal sliding. PISM is a~shallow ice sheet
      model, which implies that the balance of stresses is approximated
      based on their predominant components. The Shallow Shelf Approximation
      (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation
      (SIA) by adding velocity solutions of the two approximations
      \citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.

      Ice rheology depends on temperature and water content through an
      enthalpy formulation \citep{Aschwanden.etal.2012}. Surface air
      temperature derived from the climate forcing (Sect.~\ref{sec:atm})
      provides the upper boundary condition to the ice enthalpy model.
      Temperature is computed subglacially to a~depth of 3\,km, where it is
      conditioned by a~lower boundary geothermal heat flux of
      70\,\unit{mW\,m^{-2}}. Although this uniform value does not account
      for the high spatial geothermal variability in the region
      \citep{Blackwell.Richards.2004}, it is, on average, representative of
      available heat flow measurements. In the low-resolution simulations,
      the vertical grid consists of 31~temperature layers in the bedrock and
      up to 51~enthalpy layers in the ice sheet, corresponding to a~vertical
      resolution of 100\,m. The high-resolution simulations 61~bedrock
      layers and up to 101~ice layers with a~vertical resolution of 50\,m.

      A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
      relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the
      sliding velocity, $\vec{v}_b$ (Table~\ref{tab:params}). The yield
      stress, $\tau_c$, is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
      where cohesion, $c_0$, is assumed to be zero. The friction angle,
      $\phi$, varies from 15 to 45\degree as a~piecewise-linear function of
      modern bed elevation, with the lowest value occuring below modern sea
      level (0\,m above sea level, m~a.s.l.) and the highest value occuring
      above the generalised elevation of the highest shorelines
      \citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
      a~weakening of till associated with the presence of marine sediments
      (cf. \citealp{Martin.etal.2011};
      \citealp[supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
      the ice overburden stress, $\rho gh$, and the modelled amount of
      subglacial water, using a~formula derived from laboratory experiments
      with till extracted from the base of Ice Stream B in West Antarctica
      \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
      where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
      ratio and $C_c$ is a~measured compressibility coefficient
      (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
      from zero to $W_{max}=2$\,m, a~threshold above which additional melt
      water is assumed to drain off instantaneously. Finally, the bedrock
      topography responds to ice load following a~bedrock deformation model
      that includes local isostasy, elastic lithosphere flexure and viscous
      astenosphere deformation in an infinite half-space
      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
      value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is used for the
      astenosphere (Table~\ref{tab:params}) in accordance with the results
      from regional glacial isostatic adjustment modelling at the northern
      Cascadia subduction zone \citep{James.etal.2009}.

      Ice shelf calving is computed using a~double criterion. First, a
      physically-realistic calving flux is computed based on eigenvalues of
      the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
      Levermann.etal.2012}. This allows floating ice to advance in confined
      embayments, but prevents the formation of extensive ice shelves in the
      open ocean. Second, floating ice thinner than 50\,m is systematically
      calved off. A~subgrid scheme by \citet{Albrecht.etal.2011} allows for
      a~continuous migration of the calving front. This formulation of
      calving has been applied to the Antarctic ice sheet and has shown to
      produce a~realistic calving front positon for many of the present-day
      ice shelves \citep{Martin.etal.2011}.

@@

      The PDD computation accounts for stochastic temperature variations by
      assuming a~normal temperature distribution of standard deviation
      $\sigma$ aroung the expected value $T_m$. It is expressed by an
      error-function formulation \citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
      which is numerically approximated using week-long sub-intervals. In
      order to account for the effects of spatial and seasonal variations of
      temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
      from NARR daily temperature values from 1979 to 2000
      \citep{Mesinger.etal.2006}, including variability associated with the
      seasonal cycle (Fig.~\ref{fig:atm}). Degree-day factors for snow and
      ice melt are derived from mass-balance measurements on contemporary
      glaciers from the Coast Mountains and Rocky Mountains in British
      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

      The present-day monthly climatology was computed from near-surface air
      temperature and precipitation rate fields from the NARR, averaged from
      1979 to 2000. Modern climate of the North American Cordillera is
      characterised by strong geographic variations in temperature
      seasonality, timing of the maximum annual precipitation, and daily
      temperature variability (Fig.~\ref{fig:atm}). The use of NARR is
      motivated by the need for an accurate, high-resolution precipitation
      forcing, as identified in a~previous sensitivity study
      \citep{Seguinot.etal.2014}.

      Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
      palaeo-temperature proxy records from the Greenland Ice Core Project
      \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice Core
      Project \citep[NGRIP,][]{Andersen.etal.2004}, the European Project for
      Ice Coring in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, the
      Vostok ice core \citep{Petit.etal.1999}, and Ocean Drilling Program
      (ODP) sites 1012 and 1020, both located off the coast of California
      \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the GRIP
      and NGRIP records were calculated from oxygen isotope
      (\chem{\delta^{18}O}) measurements using a~quadratic equation
      \citep{Johnsen.etal.1995},
\begin{align}
    {\Delta}T_{TS}(t) = &-11.88 [\chem{\delta^{18}O}(t)
                                -\chem{\delta^{18}O}(0)] \nonumber \\
                        &-0.1925[\chem{\delta^{18}O}(t)^2
                                 -\chem{\delta^{18}O}(0)^2] \,,
\end{align}
      while temperature reconstructions from Antarctic and oceanic cores
      were provided as such. All records were scaled linearly
      (Table~\ref{tab:records}) in order to simulate comparable ice extents
      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
      (Fig.~\ref{fig:lr_maps}).

      Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a
      function of ice surface elevation, $s$, using the NARR surface
      geopotential height invariant field as a~reference topography,
      $b_{ref}$:
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
\end{align}

@@

      Despite large differences in the input climate forcing
      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
      features that can be observed across the range of forcing data used.
      In all simulations, modelled ice volumes remain relatively low during
      most of the glacial cycle, except during two major glacial events
      which occur between 61.9 and 56.5\,ka during MIS~4, and between 23.2
      and 16.8\,ka during MIS~2 (Fig.~\ref{fig:lr_ts}, lower panel). An ice
      volume minimum is consistently reached between 53.0 and 41.3\,ka
      during MIS~3. However, the magnitude and precise timing of these three
      events depend significantly on the choice of proxy record used to
      derive a~time-dependent climate forcing (Table~\ref{tab:extrema}).

      Simulations forced by the Greenland ice core palaeo-temperature
      records (GRIP, NGRIP) produce the highest variability in modelled ice
      volume throughout the last glacial cycle. In contrast, simulations
      driven by oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
      palaeo-temperature records generally result in lower ice volume
      variability throughout the simulation length, resulting in lower
      modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
      The NGRIP climate forcing is the only one that results in a~larger ice
      volume during MIS~4 than during MIS~2.

      While simulations driven by the GRIP and the two Antarctic
      palaeo-temperature records attain a~last ice volume maximum between
      19.1 and 16.8\,ka, those informed by the NGRIP and the two oceanic
      palaeo-temperature records attain their maximum ice volumes thousands
      of years earlier. Moreover, the ODP~1012 run yields a~rapid
      deglaciation of the modelled area prior to 17\,ka. The ODP~1020
      simulation predicts an early maximum in ice volume at 23.2\,ka,
      followed by slower deglaciation than modelled using the other
      palaeo-temperature records. Finally, whereas model runs forced by the
      Antarctic palaeo-temperature records result in a~rapid and
      uninterrupted deglaciation after the LGM, the simulation driven by the
      GRIP palaeo-temperature record also results in a~rapid deglaciation
      but in three steps, separated by two periods of ice sheet regrowth
      (Fig.~\ref{fig:lr_ts}).

@@

      In the MIS~3 ice volume minimum reconstructions, a~central ice cap
      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
      panels). Although this ice cap is present in all simulations, its
      dimensions depend sensitively on the choice of the applied
      palaeo-temperature record. Modelled ice volume minima spread over a
      wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).

      Modelled ice sheet geometries during the LGM (MIS~2;
      Fig.~\ref{fig:lr_maps}, lower panels) invariably include a~ca.
      1500\,km-long central divide above 3000\,m~a.s.l. located along the
      spine of the Rocky Mountains. Although the similarity of modelled ice
      extents is a~direct result of the choice of scaling factors applied to
      different palaeo-temperature proxy records (Table~\ref{tab:records}),
      it is interesting to note that modelled maximum ice volumes also fall
      within a~tight range of 8.40 to 8.91\,m~s.l.e.
      (Table~\ref{tab:extrema}).
